---
title: "FOG time to events analysis"
author: "Christophe Mpaga"
output:
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Executive summary
In this project, we will be using data from  [kaggle](https://www.kaggle.com/competitions/tlvmc-parkinsons-freezing-gait-prediction/data).We aim to analysis time to freezing of gait (FOG). [FOG]() is a pattern occurring in patient with Parkinson disease. It indicates kinetic inability and impairment during gait for instance. Some indicative events like walking hesitation, turning body could be observed and help to detect FOG occurence. 

# Introduction  
We aim at analyzing the time to event in Parkinsonian patients. Here FOG is the event of interest and it is expected within the execution of a given task, or during daily life activity. Here we will analyse data from lab tasks. Our main goal is to analyse and explain overall FOG time to event model, the influence/importance of some features to the occurrence of freezing of gait(FOG) from the beginning to the end of the task. To achieve this, we model our data under [kaplan-Meier]() model. 

```{r}
rm(list=ls())
```

```{r libraries,message = FALSE}
library(tidyverse)
library(skimr)
library(survival)
library(broom)
```

```{r load,  message=FALSE}
# load data from kaggle 
# from CLI kaggle competitions download -c tlvmc-parkinsons-freezing-gait-prediction 
events <- read_csv("events.csv")
subjects <- read_csv("subjects.csv")
tasks <- read_csv("tasks.csv")
#daily_meta <- read_csv("daily_metadata.csv")
defog_meta <- read_csv("defog_metadata.csv")
#tdcsfog_meta <- read_csv("tdcsfog_metadata.csv")
```

# Data preparation, inspection and preprocessing.  
let's join all metadata tables before diving into analysis. 

* defog_metadata.csv Identifies each series in the tdcsfog dataset by a unique Subject, Visit, Test, Medication condition.

    - `Visit`  Lab visits consist of a baseline assessment, two post-treatment assessments for different treatment stages, and one follow-up assessment.
    - `Test` Which of three test types was performed, with 3 the most challenging.
    - `Medication`  Subjects may have been either off or on anti-parkinsonian medication during the recording.

* subjects.csv Metadata for each Subject in the study, including their `Age` and `Sex` as well as:

    - `Visit`  Only available for subjects in the daily and defog datasets.
    - `YearsSinceDx` Years since Parkinson's diagnosis.
    - `UPDRSIIIOn/UPDRSIIIOff`  Unified Parkinson's Disease Rating Scale score    during on/off medication respectively.
  `NFOGQ` Self-report FoG questionnaire score. See:
  https://pubmed.ncbi.nlm.nih.gov/19660949/

* events.csv Metadata for each FoG event in all data series. The event times agree with the labels in the data series.  

   - `Id` The data series the event occured in.
   - `Init` Time (s) the event began.
   -  `Completion` Time (s) the event ended.
   -  `Type` Whether StartHesitation, Turn, or Walking.
   - `Kinetic` Whether the event was kinetic (1) and involved movement, or akinetic (0) and static.

* tasks.csv Task metadata for series in the defog dataset. (Not relevant for the series in the tdcsfog or daily datasets.)

  - `Id` The data series where the task was measured.
  - `Begin` Time (s) the task began.
  - `End` Time (s) the task ended.  
Task One of seven tasks types in the DeFOG protocol, described on this page.

Remove **Visit** from **subjects** table , as we do not use it 
```{r}
subjects$Visit=NULL 
```


```{r join data,warning=FALSE,message=FALSE}
  fog <- inner_join(events,tasks,by="Id") %>% 
  inner_join(defog_meta,by ="Id")  %>% 
  inner_join(subjects,by = "Subject") 
  #inner_join(tdcsfog_meta) 
  #inner_join(daily_meta)
rm(events,tasks,defog_meta,subjects)
```

Note: we are going to consider only defog condition subjects. 
  
## Data structure
Here we will refer to Kinetic as status (will change it later) and init as Time , 
then we will compute duration . 

### Ad event and tasks duration columns features: 
  - `eventsDuration <- Completion - Init `
  - `tasksDuration <- Begin - end `   
```{r add duration col to events, message=FALSE}
fog <- fog %>% filter(Begin > 0 & End > 0, End >= Completion,Begin < Init ) 

attach(fog)
# events duration
fog["eventsDuration"] <- Completion - Init


# task duration
fog["tasksDuration"] <- End - Begin

# attach fog 
detach(fog)
attach(fog)

```

# Data analysis 
## 1D EDA  
## Subjects 
How many unique subjects are in this dataset ?  

There are `r length(unique(fog$Subject))` distinct subjects in this study.  


```{r summary stats fog,eval = FALSE}
skim_without_charts(fog)
```

## Kinetic / Events 
How many trials has missing kinetic/status ?
```{r,eval = T}
round(mean(is.na(Kinetic)),2)
```

`r 100*round(mean(is.na(Kinetic)),2)`% of trials has missing Kinetic(status). 

filter for missing kinetic entries/cases.  
```{r filter row with missing kinetic values }
fog <- fog %>% 
    filter(!is.na(Kinetic))
cat("\n new table dimension\n")
dim(fog)
```

```{r,eval=TRUE}
head(fog,3)
```

```{r}
fog_grouped <- fog %>%
            group_by(Subject) 
fog_num <- fog %>% group_by(Subject) %>% 
 summarise_if(.funs = "mean",
              .predicate = is.numeric
              )
```

```{r}
dim(fog_grouped)
```

```{r}
dim(fog_num)
```

Among remaining `r 100 - 100*round(mean(is.na(Kinetic)),2)`% of trials.
What is the proportion of observed Kinetic events ?    
Kinetic graphical summary  
```{r}
barplot(table(Kinetic), main = "barplot kinetic")
```

Kinetic numerical summary  
```{r prop table kinetic events}
proportion_kinetic <- prop.table(table(Kinetic)) %>% round(2)
```
`r 100*proportion_kinetic[2]` % of events has been observed.  

Note : Given that each event is indicative of FOG, we will just gather them and consider that they form one class (Kinetic / events). 


```{r,eval=TRUE}
#How many type of kinetic are observed ?  
 barplot(table(Type))
``` 

events type numerical summary 
```{r proportion of kinetic events in table,eval=TRUE}
prop.table(table(Type)) %>% round(2)

#The most frequent events is Turn. (`r 100*prop.table(table(Type))[2] %>% round(2)`% of the observed events). 
```

events  Duration graphical summary.  
```{r,eval=TRUE}
hist(eventsDuration,freq = F,breaks = 50, xlim = c(0,60))
#abline(v = mean(eventsDuration), col ="red")
#abline(v = median(eventsDuration), col ="blue")
```
Events duration is asymetric and right skewed.  
Many events duration are of high values.   

EventsDuration Numerical summary  
```{r}
summary(eventsDuration)
```
At least 50% of cases has  `r median(eventsDuration)`s events duration, and events duration ranges between  0 .11 and 581 seconds.   

## Duration 
What is the mean duration of events, tasks and Time ?  
```{r}
cat("events mean duration :",  mean(eventsDuration),"s \n")
cat("tasks mean duration :", mean(tasksDuration),"s \n")
#cat("mean time to events duration :", mean(Time),"s \n")
```
## Age 
What is the median age ?  
Age graphical summary.  
```{r,eval=TRUE}
# age
hist(Age,breaks = 15,freq = F)
```
Age is not normally distributed, bimodal with some outliers.  
Age numerical summary.  
```{r}
summary(Age)
```

At least half of subjects are  69 years old.   

## Sex  
Sex graphical summary.  
```{r, eval = FALSE}
# sex
barplot(table(Sex))
```

Sex  numerical summary
```{r,eval=TRUE}
(sex_prop_table <- prop.table(table(Sex)) %>% round(2))
```

There is almost `r 100*prop.table(table(Sex))[2]`% of men in this cohort. 

## Tasks
What kind of tasks has been performed ?
```{r}
unique(Task)
```

how many tasks had been performed ?  
```{r,eval = FALSE}
length(unique(Task))
```

There had been `r length(unique(Task))` task performed. 

## Visit 
How many rounds of visit did the patient had ?
```{r}
length(unique(Visit))
```

What is the proportion of patient within each number of visit round ?
```{r}
round(prop.table(table(Visit)),2)
```
`r 100*round(prop.table(table(Visit))[1],2)`% of patients had one Visit, while the remaining had 2. 

## Medication 
How many subjects are under parkinson-medication ?
```{r, eval=TRUE}
100*round(prop.table(table(Medication)),4)
```

`r 100*round(prop.table(table(Medication)),4)[2]`% are under medication. 
## 
```{r}

```

## 

## 2D EDA

# Modelling and Analysis : overall time to FOG. 
```{r}
#par(mfrow = c(1,2))
#fit task duration 
fit_tti <- survfit(Surv(Begin,Init, Kinetic) ~ 1, data = fog)

# fit from begin to end
fit_tte <- survfit(Surv(Begin,End, Kinetic) ~ 1, data = fog)

```


```{r plot fit task duration}
plot(fit_tti,fun ="F",
     xlim = c(0,250),
      xlab = "Time  (seconds)",
     ylab = "Cum. Prob. of experiencing a FOG",
     main = "FOG cumul. hazard from task beginning  
     to FOG initiation",
     frame = F)
```

```{r fit tti summary}
fit_tti
```

When considering time to FOG initiation , half of patients experiment FOG within `r summary(fit_tti)$table["median"]` s. 

```{r plot fit begin to end }
# plot fit2
plot(fit_tte,
     fun ="F",
     xlim =c(0,250),
      xlab = "Time (seconds)",
     ylab = "Cum. Prob. of experiencing a FOG", 
     main ="FOG cum. hazard from task start 
     to the end",
     frame = F)
```

```{r fit_tte summary}
fit_tte
```
When considering time to the end of the task, half of patients experiment FOG within `r summary(fit_tte)$table["median"]` s. 

## group effect : coxph analysis. 
### sex
```{r}
# fit
fit_sex <- coxph(Surv(Begin,End, Kinetic) ~ factor(Sex), data = fog)

# summary fit 
p <- round(tidy(fit_sex)$p.value,2)
```
p-value is `r p` .There is no effect of sex on time to FOG. 

### Medication
```{r}
fit_Med <- coxph(Surv(Begin,End, Kinetic) ~ factor(Medication), data = fog)

# summary fit_medication
p <- round(tidy(fit_Med)$p.value,2)
```
P-value is `r p`<< 0.05. there is huge effect of Medication in experiencing FOG holding other variables fixed. 

### effect of task type 
```{r}
fit_task <- coxph(Surv(Begin,End, Kinetic) ~ factor(Task), data = fog)

# summary type 
p <- round(tidy(fit_task)$p.value,2)

```
p-value `r p` is << to 0.05, there is strong effect of task type in FOG occurence holding other variables fixed. 

### effect of event type 
```{r }
fit_type <- coxph(Surv(Begin,End, Kinetic) ~ Task, data = fog)

# summary fit_type
tidy(fit_type)
```
p-value is `r p` is << to 0.05. There an effect of task type in the risk of experiencing FOG. 

### effect of visit 
```{r}
fit_visit <- coxph(Surv(Begin,End, Kinetic) ~ factor(Visit), data = fog)

# summary fit_type
p <- round(tidy(fit_visit)$p.value,2)
```
P-value is `r p` and is << to 0.05, there is an effect of visit number in FOG occurrence.  

# Discussion and Conclusion


# References 


**To be continued !!!!**



