---
title: "Time to freezing of gait analysis : A time to event analysis"
output:
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


# Executive summary
In this project, we will be using data from  [kaggle](https://www.kaggle.com/competitions/tlvmc-parkinsons-freezing-gait-prediction/data).We aim at analyzing the expected duration of time until freezing of gait (FOG) occurs (time to FOG). [FOG]() is a pattern occurring in patient with Parkinson disease. It is a debilitating symptom that afflicts many people with Parkinson’s disease. Some events like start walking hesitation, turning body could be observed and help to detect FOG occurrence. 

# Introduction  
We aim at analyzing the time to FOG in 44 Parkinson disease patients, with median age of 69 years old. Here FOG is the event of interest and it is expected within the execution of a given task in lab, or during daily life activity.We will analyse data from lab tasks. Our main goal is to analyse and explain overall FOG time to event model, the influence/importance of some features in the time to occurrence of FOG from the beginning to the end of the task. To achieve this, we model our data under [kaplan-Meier]() model. 

```{r}
rm(list=ls())
```

```{r libraries,message = FALSE}
library(tidyverse)
library(survival)
library(broom)
library(survminer)
```

```{r load,  message=FALSE}
# load data from kaggle 
# from CLI kaggle competitions download -c tlvmc-parkinsons-freezing-gait-prediction 
events <- read_csv("events.csv")
subjects <- read_csv("subjects.csv")
tasks <- read_csv("tasks.csv")
#daily_meta <- read_csv("daily_metadata.csv")
defog_meta <- read_csv("defog_metadata.csv")
#tdcsfog_meta <- read_csv("tdcsfog_metadata.csv")
```

# Data preparation, inspection and preprocessing.  
let's join all metadata tables before diving into analysis. 

* defog_metadata.csv Identifies each series in the tdcsfog dataset by a unique Subject, Visit, Test, Medication condition.

    - `Visit`  Lab visits consist of a baseline assessment, two post-treatment assessments for different treatment stages, and one follow-up assessment.
    - `Test` Which of three test types was performed, with 3 the most challenging.
    - `Medication`  Subjects may have been either off or on anti-parkinsonian medication during the recording.

* subjects.csv Metadata for each Subject in the study, including their `Age` and `Sex` as well as:

    - `Visit`  Only available for subjects in the daily and defog datasets.
    - `YearsSinceDx` Years since Parkinson's diagnosis.
    - `UPDRSIIIOn/UPDRSIIIOff`  Unified Parkinson's Disease Rating Scale score    during on/off medication respectively.
  `NFOGQ` Self-report FoG questionnaire score. See:
  https://pubmed.ncbi.nlm.nih.gov/19660949/

* events.csv Metadata for each FoG event in all data series. The event times agree with the labels in the data series.  

   - `Id` The data series the event occured in.
   - `Init` Time (s) the event began.
   -  `Completion` Time (s) the event ended.
   -  `Type` Whether StartHesitation, Turn, or Walking.
   - `Kinetic` Whether the event was kinetic (1) and involved movement, or akinetic (0) and static.

* tasks.csv Task metadata for series in the defog dataset. (Not relevant for the series in the tdcsfog or daily datasets.)

  - `Id` The data series where the task was measured.
  - `Begin` Time (s) the task began.
  - `End` Time (s) the task ended.  
Task One of seven tasks types in the DeFOG protocol, described on this [page](https://www.kaggle.com/competitions/tlvmc-parkinsons-freezing-gait-prediction/data).

we removed **Visit** from **subjects** table , as we do not use it 
```{r}
subjects$Visit=NULL 
```


```{r join data,warning=FALSE,message=FALSE}
  fog <- inner_join(events,tasks,by="Id") %>% 
  inner_join(defog_meta,by ="Id")  %>% 
  inner_join(subjects,by = "Subject") 
rm(events,tasks,defog_meta,subjects)
```

Note: we are going to consider only defog condition subjects. 
  
## Data structure
### Ad event and tasks duration columns features: 
  - `eventsDuration <- Completion - Init `
  - `tasksDuration <- Begin - end `   
```{r add duration col to events, message=FALSE}
fog <- fog %>% filter(Begin > 0 & End > 0, End >= Completion,Begin < Init ) 

# change some variables into factor 
fog$Sex <- factor(fog$Sex, labels =c("Female","Male") )
fog$Type <- factor(fog$Type)
fog$Task <- factor(fog$Task)
fog$Medication <- factor(fog$Medication)
fog$Visit <- factor(fog$Visit)

attach(fog)
# events duration
fog["eventsDuration"] <- Completion - Init


# task duration
fog["tasksDuration"] <- End - Begin

# attach fog 
detach(fog)
attach(fog)


```

# Data analysis 
## Subjects 
How many unique subjects are in this dataset ?  

There are `r length(unique(fog$Subject))` distinct subjects in this study.  


```{r summary stats fog,eval = FALSE}
skim_without_charts(fog)
```

## Kinetic / Events 
How many trials has missing kinetic/status ?
```{r,eval = T}
round(mean(is.na(Kinetic)),2)
```

`r 100*round(mean(is.na(Kinetic)),2)`% of trials has missing Kinetic(status). 

filter for missing kinetic entries/cases.  
```{r filter row with missing kinetic values }
fog <- fog %>% 
    filter(!is.na(Kinetic))
cat("\n new table dimension\n")
dim(fog)
```

```{r}
head(fog,3)
```

What is the proportion of observed Kinetic events ?    
```{r prop table kinetic events}
proportion_kinetic <- prop.table(table(Kinetic)) %>% round(2)
```
`r 100*proportion_kinetic[2]` % of events has been observed.  

Note : Given that each event is indicative of FOG, we will just gather them and consider that they form one class (Kinetic / events). 

## Age 
What is the median age ?  
Age numerical summary.  
```{r}
summary(Age)
```
At least half of subjects are  69 years old or above.   

## Sex  
Sex  numerical summary
```{r,eval=TRUE}
(sex_prop_table <- prop.table(table(Sex)) %>% round(2))
```

There is almost `r 100*prop.table(table(Sex))[2]`% of men in this cohort. 

## Tasks
What kind of tasks has been performed ?
```{r}
unique(Task)
```

how many tasks had been performed ?  
```{r,eval = FALSE}
length(unique(Task))
```

There had been `r length(unique(Task))` task performed. 

## Visit 
How many rounds of visit did the patient had ?
```{r}
length(unique(Visit))
```

What is the proportion of patient within each number of visit round ?
```{r}
round(prop.table(table(Visit)),2)
```
`r 100*round(prop.table(table(Visit))[1],2)`% of patients had one Visit, while the remaining had 2. 

## Medication 
How many subjects are under parkinsonian-medication ?
```{r, eval=TRUE}
100*round(prop.table(table(Medication)),4)
```

`r 100*round(prop.table(table(Medication)),4)[2]`% are under medication. 

# Modelling and Analysis : overall time to FOG. 
```{r}
#par(mfrow = c(1,2))
#fit task duration 
fit_tti <- survfit(Surv(tasksDuration, Kinetic) ~ 1, data = fog)

# fit from begin to end
fit_tte <- survfit(Surv(Begin,End, Kinetic) ~ 1, data = fog)

```


```{r plot fit time to initiation,eval=FALSE}
plot(fit_tti,
     fun ="F",
     xlim = c(0,250),
      xlab = "Time  (seconds)",
     ylab = "Cum. Prob. of experiencing a FOG",
     main = "FOG cumul. prob from task beginning  
     to FOG initiation",
     frame = F)
```

```{r fit tti summary,eval = F}
fit_tti
When considering time to FOG initiation , half of patients experiment FOG within `r summary(fit_tti)$table["median"]` s. 
```



```{r plot fit begin to end }
# plot fit2
plot(fit_tte,
     fun ="F",
     xlim =c(0,250),
      xlab = "Time (seconds)",
     ylab = "Cum. Prob. of experiencing a FOG", 
     main ="FOG cum. prob has median 28.9 s",
     frame = F)
abline(v = 28.9, col = "red")
```

```{r fit_tte summary}
fit_tte
```
Half of patients experience FOG within `r summary(fit_tte)$table["median"]` s. 

## Univariate analysis : Survival difference in group
### Sex
```{r}
# fit
(fit_sex <- survfit(Surv(Begin,End, Kinetic) ~ Sex, data = fog))

```

```{r}
# coxph
coxph_sex <- coxph(Surv(Begin,End, Kinetic) ~ Sex, data = fog)
p <- round(tidy(coxph_sex)$p.value,2)
```
p-value is `r p` .There is no effect of sex on time to FOG. 

### Medication
```{r}
fit_Med <- survfit(Surv(Begin,End, Kinetic) ~ Medication, data = fog)
coxph_Med <- coxph(Surv(Begin,End, Kinetic) ~ Medication, data = fog)

# summary fit_medication
p <- round(tidy(coxph_Med)$p.value,2)
```
P-value is `r p` which is less than 0.05. There is a huge effect of Medication in experiencing FOG holding other variables fixed. 

```{r}
fit_Med
```

```{r}
ggsurvplot(fit_Med, 
     fun = "event",
     #color = "strata",
     conf.int = F,
     risk.table.col="strata",
     surv.median.line = "hv",
     size = 0.7,
     xlim = c(0,300),
     break.time.by = 50,
     xlab = "Time (seconds)",
     ylab = "cum. prob. time to FOG",
     title = "Time to fog is short for subjects under medication")

```


### Task type 
```{r}
fit_task <- survfit(Surv(Begin,End, Kinetic) ~ Task, data = fog)
coxph_task <- coxph(Surv(Begin,End, Kinetic) ~ Task, data = fog)

# summary type 
tidy(coxph_task) |> 
  filter(p.value <= 0.05) |>
  select(term, estimate,p.value)
```

Three levels have significant effect on time to FOG,holding other variables fixed.  

### events Type 
```{r }
fit_Type <- survfit(Surv(Begin,End, Kinetic) ~ Type, data = fog)
coxph_Type <- coxph(Surv(Begin,End, Kinetic) ~ Type, data = fog)

# summary fit_type 
tidy(coxph_Type) |> 
  filter(p.value <= 0.05) |>
  select(term, estimate,p.value)
```

```{r}
ggsurvplot(fit_Type,
     fun ="event",
     xlim =c(0,250),
     break.time.by = 50,
     surv.median.line = "hv",
     size = 0.7,
     xlab="Time(seconds)",
     ylab = "cum prob of time to FOG w/r to type events"
     )
```

2  p.values are greater than 0.05. There is no difference of Time to events in FOG occurrence.

### visit 
```{r}
(fit_visit <- survfit(Surv(Begin,End, Kinetic) ~ Visit, data = fog))
coxph_visit <- coxph(Surv(Begin,End, Kinetic) ~ Visit, data = fog)

# summary fit_type
tidy(coxph_visit) |> 
  filter(p.value <= 0.05) |>
  select(term, estimate,p.value)
```

```{r}
ggsurvplot(fit_visit,
     fun ="event",
     conf.int = F,
     risk.table.col="strata",
     surv.median.line = "hv",
     size = 0.7,
     break.time.by = 50,
     xlim = c(0,250),
    xlab ="Time(seconds)",
     ylab ="cum prob of time to FOG w/r to visit to hospital")
```

p-value is less than  0.05. Having visit to hospital, has an influence on FOG occurrence.  


### yearSinceDx : year since diagnosis.  
```{r}
(fit_YSDx<- coxph(Surv(Begin,End, Kinetic) ~ YearsSinceDx,
                 data = fog))
coxph_YSDx<- coxph(Surv(Begin,End, Kinetic) ~ YearsSinceDx,
                 data = fog)
p <- round(tidy(coxph_YSDx)$p.value,2)
```
p-value is `r p` which is > 0.05. There is no effect of Year since diagnosis on FOG occurrence.  

## Multivariate analysis 
```{r}
fit_mult<- coxph(Surv(Begin,End, Kinetic) ~ (Sex + Age +  Medication  +
                                               Visit + YearsSinceDx),
                 data = fog)
tidy(fit_mult) %>% 
  filter(p.value <= 0.05)
```


Medication and Visit present a p value < 0.05. Medication increases the risk of experiencing FOG by 0.7 holding visit fixed, while , visit does the same with an increase of time to FOG by a rate of 0.2 while holding other features fixed. 

# Results  
This cohort has 44 subjects, half of the subjects are at least 69 years old. `r 100*round(prop.table(table(Medication))[2],4)` % are under medication,  `r  100*round(prop.table(table(Visit))[2],2)`% present more than 1 round visit to hospital. Our analysis shows that at least half of the subjects time to FOG is  `r summary(fit_tte)$table["median"]` seconds. Subjects  under medication and having frequent visit to hospital are more likely to experience a FOG episode during a motor/kinetic  task like gait. 

# Conclusion   
We aim at analyzing time to FOG in this study. Our analysis shows that FOG occurrence is related to Medication and Visit to hospital. Overall, half of subjects experience FOG within 28.9 seconds.It could be interesting to investigate also survival difference given kinetic outcomes and tasks. 

# References 
Cox, David R. 1972. “Regression Models and Life-Tables.” Journal of the Royal Statistical Society: Series B (Methodological) 34 (2): 187–202.  
Kaplan, Edward L, and Paul Meier. 1958. “Nonparametric Estimation from Incomplete Observations.” Journal of the American Statistical Association 53 (282): 457–81.

```{r}

```




