---
title: "FOG"
author: "Christophe"

output: 
    html_document:
        toc: true
        toc_float : true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Executive summary
In this project, we will be using data from  [kaggle](https://www.kaggle.com/competitions/tlvmc-parkinsons-freezing-gait-prediction/data).In this data, we will try to analysis time to freezing of gait (FROG). FOG is a pattern occurring in patient with Parkinson diseases. It is characterize by kinetic inability and impairment. some indicative events like walking hesitation, turning body could be observed. 

```{r libraries,message = FALSE}
library(tidyverse)
library(skimr)
```

```{r load}
# load data from kaggle 
# from CLI kaggle competitions download -c tlvmc-parkinsons-freezing-gait-prediction 
events <- read_csv("events.csv")
subjects <- read_csv("subjects.csv")
tasks <- read_csv("tasks.csv")
```
# events 
````{r}
str(events)
```


```{r}
summary(events)
```
Kinetic has 1042 NA.  

```{r Kinetic plot}
barplot(table(events$Kinetic))
```
Most kinetic event status is "1", i.e. many events occurs. 


```{r}
prop.table(table(events$Kinetic)) %>% round(2)
```

82 % of events has been observed during this experiment. 

```{r proportion of kinetic events in table}
prop.table(table(events$Type)) %>% round(2)
```

```{r}
 with(events,barplot(table(Type)))
```


The most frequent events is Turn. (80% of the observed events).  
```{r add duration col to events}
events["eventsDuration"] <- with(events,Completion - Init)
head(events,3)
```

```{r}
with(events,hist(eventsDuration,freq = F,breaks = 100, xlim = c(0,60)))
```

EventsDuration is asymetric , with median 2.7 Unit time. 
```{r}
with(events,summary(events_duration))
```
== subjects ==

```{r}
str(subjects)
```
```{r}
dim(subjects)
```

```{r}
dim(unique(subjects)) 
```

There is 173 unique subjects.  
```{r}
# age
with(subjects,hist(Age,breaks = 25))
```

```{r}
with(subjects,summary(Age))
```
There is 94 NA in Age.  Age has median of 68 years old and presents some outliers. 

```{r}
# sex
with(subjects,barplot(table(Sex)))
```

```{r}
with(subjects,prop.table(table(Sex))) %>% round(2)
```

There is almost 70% of men in this cohort. 

```{r}
str(tasks)
```



```{r}
```



```{r}
```

